Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2018-05-28 survey/survey.tex
sec 2

Some of the simplest spaces are the finite and discrete spaces of
points. Indeed most programming languages provide some collection of
primitive data like booleans, characters, strings, and (bounded)
numbers that are naturally modeled as finite sets. Infinite sets are much
more subtle, and will be discussed in the conclusion.

@@

Although this language appears weak, it is universal for boolean
combinational circuits. The result follows from the following
construction of the Toffoli gate and the fact that any combinational
circuit can be realized using compositions of Toffoli gates. We write
$[2^n]$ for the finite set representing $n$-bit words with the
canonical ordering for binary numbers. Thus when $n=3$, the finite set
has elements $\{0,1,2,3,4,5,6,7\}$ which correspond to the 3-bit words
$\{000,001,010,011,100,101,110,111\}$. The Toffoli gate is a
reversible function $[2^3] \rightarrow [2^3]$ which swaps $110$ and
$111$, i.e., it can be realized as $\mathit{swap}~6~7$.

\amr{let's write a a couple more small examples, reversible 3 bit
  adder, etc.}

In programming practice we are however more interested in structured
data which will be the subject on next section. What we learn from
this short investigation using untyped flat sets is what the
``\emph{purely operational}'' view of the theory would be. In
particular, it tells us that permutations are an inescapable part of
the fabric of reversible computing.  However as permutations are
untyped, and act on the canonicalized version of $n$-element sets
(i.e. those sets where all the structure has been forgotten), these
are a rather pale shadow of the rich tapestry of
information-perserving transformations of structured data which we
investigate next.

>>>>

Most programming languages provide primitive data like booleans,
characters, strings, and (bounded) numbers that are naturally modeled
as finite sets. We therefore start by modeling reversible computations
over finite and discrete spaces of points. Infinite sets are more
subtle, and will be discussed in the conclusion.

@@

For convenience, we write $[2^n]$ for the finite set representing
$n$-bit words with the canonical ordering for binary numbers. Thus
when $n=3$, the finite set has elements $\{0,1,2,3,4,5,6,7\}$ which
correspond to the 3-bit words $\{000,001,010,011,100,101,110,111\}$.
Although this language appears weak, it is universal for boolean
combinational circuits. To see this, we express, boolean negation
\emph{not} as a permutation $[2^1] \rightarrow [2^1]$, controlled not
(also known as \emph{cnot}) as a permutation
$[2^2] \rightarrow [2^2]$, and controlled controlled-not (also known
as \emph{toffoli}) as a permutation $[2^3] \rightarrow [2^3]$:
\[\begin{array}{rcl}
\mathit{not} &=& \mathit{swap}~0~1 \\
\mathit{cnot} &=& \mathit{swap}~2~3 \\
\mathit{toffoli} &=& \mathit{swap}~6~7
\end{array}\]
For example, \emph{toffoli} gate negates the third bit (the target
bit) if both the first two bits (the control bits) are 1, i.e., it
swaps $110$ and $111$. Universality follows since any combinational
circuit can be realized using compositions of \emph{toffoli} gates.

There is however a subtle issue: programming in such an unstructured
language is \emph{not} compositional in the sense that using
$\mathit{not}$ in a larger circuit forces us to change its
implementation. To illustrate this point, consider the reversible full
adder below designed by Desoete et al.~\cite{117414}:

\begin{center}
\includegraphics[scale=0.6]{full-adder.png}
\end{center}

In the figure (copied from a more general paper that includes
alternative designs~\cite{Rentergem2005OptimalDO}), the full adder
takes 4 inputs: the two bits to add $A$ and $B$, an incoming carry bit
$C_i$, and a heap input initialized to 0 to maintain
reversibility. There are four outputs: the first two are identical to
the incoming bits $A$ and $B$ and are considered ``garbage.'' The
third bit $S$ is the sum and the last bit $C_o$ is the outgoing carry
bit. In the notation used to describe the circuit, the $x$ denotes
boolean negation, the dots are control bits. In our reversible
language, we can express this circuit as the following permutation of
type $[2^4] \rightarrow [2^4]$:

\[\begin{array}{ll}
\mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{toffoli} \\
\mathit{swap}~8~12 \fatsemi \mathit{swap}~9~14 \fatsemi 
    \mathit{swap}~10~13 \fatsemi \mathit{swap}~11~15 & \mbox{cnot and swap} \\
\mathit{swap}~6~7 \fatsemi \mathit{swap}~14~15 & \mbox{toffoli} \\
\mathit{swap}~4~6 \fatsemi \mathit{swap}~5~7 \fatsemi
    \mathit{swap}~12~14 \fatsemi \mathit{swap}~13~15 & \mbox{cnot}
\end{array}\]

Note how the implementation of $\mathit{cnot}$ as a permutation
$[2^2] \rightarrow [2^2]$ cannot be directly reused in a circuit
$[2^4] \rightarrow [2^4]$. Indeed, in programming practice we are
interested in structured data and compositional abstractions which
will be the subject on next section. What we do learn however from
this short investigation using untyped flat sets is what the
``\emph{purely operational}'' view of the theory would be. In
particular, it tells us that permutations are an inescapable part of
the fabric of reversible computing.  However as permutations are
untyped, and act on the canonicalized version of $n$-element sets
(i.e. those sets where all the structure has been forgotten), these
are a rather pale shadow of the rich tapestry of
information-perserving transformations of structured data which we
investigate next.
